---
title: "Final Project"
author: "Cesar L. Espitia"
date: "12/9/2018"
output:
  html_document: default
  pdf_document: default
---

## Statement of the Problem or Need

Education is a very important topic in any society.    One of the unique aspects of the US education system is the levels of diversity in our public-school systems.    Over the past decade, research at various levels in the public and private sector have determined that diversity is important to success and it all starts with a person’s education.  

The goal of this project is to determine the effects of a school’s student demographics on SAT scores.    This will be accomplished by using two datasets:
  •2011 -2016 Demographic Snapshot
  •2012 SAT Results

Both datasets come from the NYC OpenData portal (data.cityofnewyork.us/).  


### Importing the Data

```{r import}
library(e1071)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(VIM)
library(stringr)
library(knitr)
## Load demographic data (set 1)
demosnapshot = read.csv("2011_-_2016_Demographic_Snapshot.csv", header = TRUE)

names <- colnames(demosnapshot)

names <- sub(".1","Perc",names)

demosnapshot <- setNames(demosnapshot, names)

#convert all columns to numeric
for (i in 4:38)
{
  demosnapshot[,i] <- gsub("%", "", demosnapshot[,i])
  demosnapshot[,i] <- as.numeric(as.character(demosnapshot[,i]))
}


#lapply(demosnapshot, class)

## Load SAT Scores 2012
SATonly = read.csv("2012_SAT_Results.csv", header = TRUE)

#lapply(SATonly, class)

## Add total SAT score column for linear regression

SATonly$Math <- as.numeric(as.character(SATonly$SAT.Math.Avg..Score))
SATonly$Writing <- as.numeric(as.character(SATonly$SAT.Writing.Avg..Score))
SATonly$Reading <- as.numeric(as.character(SATonly$SAT.Critical.Reading.Avg..Score))

SATonly$SATTotal <- SATonly$Math + SATonly$Writing + SATonly$Reading

SATonly <- SATonly[ , -which(names(SATonly) %in% c("SAT.Math.Avg..Score","SAT.Writing.Avg..Score","SAT.Critical.Reading.Avg..Score"))]

## The assumption is that 2012 corresponds to the school year 2012-13 in demosnapshot
SATonly$Year <- '2012-13'

## total test takers has s to signify no one

SATonly$Num.of.SAT.Test.Takers[
SATonly$Num.of.SAT.Test.Takers == "s"] <- "0"

SATonly[,3] <- as.numeric(as.character(SATonly[,3]))

df <- merge(demosnapshot,SATonly,by=c("DBN","Year"), all=T)
df <- df[ , -which(names(df) %in% c("SCHOOL.NAME"))]

```


### Exploratory Analysis

```{r exploratory}
df2012 <- merge(demosnapshot,SATonly,by=c("DBN","Year"))
  
df2012 %>%
  keep(is.numeric) %>%                     # Keep only numeric columns
  gather() %>%                             # Convert to key-value pairs
  ggplot(aes(value)) +                     # Plot the values
    facet_wrap(~ key, scales = "free") +   # In separate panels
    geom_density()  

options(scipen = 999)
missing_plot <- VIM::aggr(df2012,  
                      numbers = T, 
                      sortVars = T,
                      col = c("lightgreen", "darkred", "orange"),
                      labels=names(df2012), 
                      ylab=c("Missing Value Counts"
                             , "Pattern"))

summary(missing_plot)

missing_plot$missings %>% 
  mutate(
    pct_missing = Count / nrow(df2012)
    ) %>% 
  arrange(-pct_missing) %>% 
  filter(pct_missing > 0) %>% 
  kable(digits = 3, row.names = T, caption = "Variables Missing Values")  

## There are 5 variables that have missing data.  This equates to 10% of the data and is acceptable.  
```


### Linear Regression

```{r linear}
# Preparing
set.seed(68)

df2 <- df2012[ , -which(names(df2012) %in% c("School.Name","SCHOOL.NAME"))]

df2 <- select_if(df2, is.numeric)

df2$Num.of.SAT.Test.Takers[is.na(df2$Num.of.SAT.Test.Takers)] <- mean(df2$Num.of.SAT.Test.Takers, na.rm=TRUE)
df2$Math[is.na(df2$Math)] <- mean(df2$Math, na.rm=TRUE)
df2$Writing[is.na(df2$Writing)] <- mean(df2$Writing, na.rm=TRUE)
df2$Reading[is.na(df2$Reading)] <- mean(df2$Reading, na.rm=TRUE)
df2$SATTotal[is.na(df2$SATTotal)] <- mean(df2$SATTotal, na.rm=TRUE)

train<-sample_frac(df2, 0.7)
sid<-as.numeric(rownames(train)) # because rownames() returns character
test<-df2[-sid,]

## Load SAT Scores 2012
#excluding math, writing, reading as these values are irrevelant.  
model1 <- lm(SATTotal~ .  - Math - Writing - Reading,data=train)

summary(model1)

test$pred1 <- predict(model1, type="response", test)

par(mfrow=c(1,2))
hist(test$SATTotal)
hist(test$pred1)

par(mfrow=c(1,2))
boxplot(test$SATTotal, data=test)
boxplot(test$pred1, data=test)


```


