---
title: "Final Project"
author:
- Cesar L. Espitia
- Duubar Villalobos Jimenez

date: "December 09, 2018"
output:
  pdf_document:
      highlight: tango
      toc: true
      toc_depth: 4
      number_sections: false
      df_print: kable
      
      
  html_document:
      df_print: paged
      code_folding: hide
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
    toc: yes
    df_print: paged
subtitle: CUNY MSDS DATA 609
fontsize: 10pt
geometry: margin=1in
always_allow_html: yes
---

```{r, echo=FALSE, warning=FALSE, error=FALSE, cache=FALSE, results='hide', message=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```


```{r, echo=FALSE, warning=FALSE, error=FALSE, cache=FALSE, results='hide', message=FALSE}
# In this section we will include the needed libraries
library(dplyr)
library(stringr)
library(data.table)

```






\newpage
# INTRODUCTION

## TITLE

**Student demographic and enrollment data from the NYC Department of Education 2011 – 2016.**

## ABSTRACT

This is our final project for DATA609. The purpose of this project is to explore, analyze and model a real-world data set of our own interest using the regression modeling techniques learned in the course. The idea is to develop a problem statement and research question based on the data set that we have obtained from the Department of Education (DOE). We will perform state-of-the-art literature and research developments dealing with empirical studies, algorithms, or methodologies related to our problem.

## DESCRIPTION

Education is a very important topic in any society. One of the unique aspects of the US education system is the levels of diversity in our public-school systems. Over the past decade, research at various levels in the public and private sector have determined that diversity is important to success and it all starts with a person’s education.  

## GOAL

The goal of this project is to determine the effects of a school’s student demographics on SAT scores. This will be accomplished by using two datasets:

    • 2011 - 2016 Demographic Snapshot
    • 2012 SAT Results

Both datasets come from the NYC OpenData portal (data.cityofnewyork.us/).  

### Modeling using Multiple Linear Regression

For this analysis, we propose to create a mathematical model to predict SAT scores based upon the school's demographics based on the 2012 data.  

### Modeling using Graph Theory

In this case, we will use the power of graphs in order to describe the relationships in between variables.

\newpage
# ANALYSIS

In this section we will incorprate the results and findings from our study analysis.

## ANALYSIS I: Modeling using Multiple Linear Regression

Cesar's work goes here...










## ANALYSIS II: Modeling with Difference Equations

In this section, we will employ a powerful modeling technique in order to have an approximation of the 2016-17 School year by employing Difference Equations we could find valuable insights.


```{r, echo=FALSE}

# Need to read datasets
url <- "https://raw.githubusercontent.com/cspitmit03/CourseProject/master/2011_-_2016_Demographic_Snapshot.csv"
demosnapshot <- read.csv(url, header = TRUE, stringsAsFactors = FALSE)

url <- "https://raw.githubusercontent.com/cspitmit03/CourseProject/master/2012_SAT_Results.csv"
SAT <- read.csv(url, header = TRUE, stringsAsFactors = FALSE)
SAT <- SAT[,-2]

# Totals have s to signify no one
SAT$Num.of.SAT.Test.Takers[SAT$Num.of.SAT.Test.Takers == "s"] <- NA
SAT$SAT.Critical.Reading.Avg..Score[SAT$SAT.Critical.Reading.Avg..Score == "s"] <- NA
SAT$SAT.Math.Avg..Score[SAT$SAT.Math.Avg..Score == "s"] <- NA
SAT$SAT.Writing.Avg..Score[SAT$SAT.Writing.Avg..Score == "s"] <- NA

# Need to convert to numeric
SAT$Num.of.SAT.Test.Takers <- as.numeric(SAT$Num.of.SAT.Test.Takers)
SAT$SAT.Critical.Reading.Avg..Score <- as.numeric(SAT$SAT.Critical.Reading.Avg..Score)
SAT$SAT.Writing.Avg..Score <- as.numeric(SAT$SAT.Writing.Avg..Score)
SAT$SAT.Math.Avg..Score <- as.numeric(SAT$SAT.Math.Avg..Score)

# Calculating SAT Total
SAT$SATTotal <- SAT$SAT.Critical.Reading.Avg..Score + SAT$SAT.Writing.Avg..Score + SAT$SAT.Math.Avg..Score

## The assumption is that 2012 corresponds to the school year 2012-13 in demosnapshot
SAT$Year <- '2012-13'

# Need complete cases only
SAT <- SAT[complete.cases(SAT),]
demosnapshot <- demosnapshot[complete.cases(demosnapshot), ]
```

```{r}
# Need to reduce data set of demsnapshot in order to employ percentages only
demosnapshot <- demosnapshot %>% 
                                select( "DBN", "School.Name", "Year", "Total.Enrollment",
                                 "X..Female.1", "X..Male.1", "X..Asian.1",
                                 "X..Black.1", "X..Hispanic.1", "X..Other.1",
                                 "X..White.1", "X..Students.with.Disabilities.1", "X..English.Language.Learners.1",
                                 "X..Poverty.1")

```




```{r, echo=FALSE}

# Function that transform Factor variables to Numeric
factor_to_numeric <- function(df){
  
  # Transform percentage to numeric.
  for (i in 5:14){
    df[,i] <- as.numeric(gsub("%", "", df[,i]))
  }
 
  return(df)
}

```

```{r, echo=FALSE}

# Need to transform from % to numeric
demosnapshot <- factor_to_numeric(demosnapshot)
 
```




```{r}

# Need to merge data in order to create data-set for linear model
lm.data <- merge(demosnapshot, SAT, 
                  by.x = c("DBN","Year"),
                  by.y = c("DBN","Year"),
                  all.y=TRUE)

lm.data
```

```{r}

# Complete cases of records to be used in order to create a linear model in order to predict the SAT score from Demographics
lm.data <- lm.data[complete.cases(lm.data),]
lm.data
```



```{r}
# From above, we have two data sets.

# -snapshot- to be used in order to obtain a model for 2016-17
# -lm.data- to be used in order to creat a linear model to predict the SAT score from the demographics
```

```{r}
# Defining data-set for linear model
lm.data1 <- lm.data[4:18]

# NEed to convert to numeric
for (i in 1:15){
  lm.data1[,i] <- as.numeric(lm.data1[,i])
}
```

```{r}
# 1) Creating linear model for SAT.Critical.Reading.Avg..Score
Rglm.SAT_NULL <- glm(SAT.Critical.Reading.Avg..Score ~ 1 
              - SAT.Math.Avg..Score - SAT.Writing.Avg..Score - Num.of.SAT.Test.Takers,
              family = "poisson", data = lm.data1)

Rglm.SAT_FULL <- glm(SAT.Critical.Reading.Avg..Score ~ . 
              - SAT.Math.Avg..Score - SAT.Writing.Avg..Score - Num.of.SAT.Test.Takers,
              family = "poisson", data = lm.data1)

Rglm.SAT_STEP <- step(Rglm.SAT_NULL,
                    scope = list(upper=Rglm.SAT_FULL),
                    direction="both",
                    test="Chisq",
                    data=lm.data1)

summary(Rglm.SAT_STEP)

# 2) Creating linear model for SAT.Writing.Avg..Score
Wglm.SAT_NULL <- glm(SAT.Writing.Avg..Score ~ 1 
              - SAT.Critical.Reading.Avg..Score - SAT.Math.Avg..Score - Num.of.SAT.Test.Takers,
              family = "poisson", data = lm.data1)

Wglm.SAT_FULL <- glm(SAT.Writing.Avg..Score ~ . 
              - SAT.Critical.Reading.Avg..Score - SAT.Math.Avg..Score - Num.of.SAT.Test.Takers,
              family = "poisson", data = lm.data1)

Wglm.SAT_STEP <- step(Wglm.SAT_NULL,
                    scope = list(upper=Wglm.SAT_FULL),
                    direction="both",
                    test="Chisq",
                    data=lm.data1)

summary(Wglm.SAT_STEP)


# 3) Creating linear model for SAT.Math.Avg..Score
Mglm.SAT_NULL <- glm(SAT.Math.Avg..Score ~ 1 
              - SAT.Critical.Reading.Avg..Score - SAT.Writing.Avg..Score - Num.of.SAT.Test.Takers,
              family = "poisson", data = lm.data1)

Mglm.SAT_FULL <- glm(SAT.Math.Avg..Score ~ . 
              - SAT.Critical.Reading.Avg..Score - SAT.Writing.Avg..Score - Num.of.SAT.Test.Takers,
              family = "poisson", data = lm.data1)

Mglm.SAT_STEP <- step(Mglm.SAT_NULL,
                    scope = list(upper=Mglm.SAT_FULL),
                    direction="both",
                    test="Chisq",
                    data=lm.data1)

summary(Mglm.SAT_STEP)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
par(mfrow=c(2,2))
plot(Wglm.SAT_STEP)
```



 This is Duubar's Work

```{r}
# First let's get the number of years provided in the data for each school
demosnapshot_nYears <- demosnapshot  %>% group_by(DBN) %>% summarise(n = n())
demosnapshot_nYears
```

```{r}
# Let's visualize it
barplot(prop.table(table(demosnapshot_nYears$n)),
        main = "Train Data-Set",
        ylab = "Probability",
        xlab = "Provided Years")
```
```{r}
# Let's see the counts
demosnapshot_nYears %>% group_by("nYears" = n) %>% summarize(count=n())
```

```{r}
# Second let's focus on the schools that have 5 years of records in the data
demosnapshot_5Years <- demosnapshot_nYears[demosnapshot_nYears$n == 5,]
# Find duplicates
#demosnapshot_5Years[duplicated(demosnapshot_5Years$DBN),]
```

```{r}
demosnapshot_SAT <- merge(demosnapshot_5Years, SAT, 
                  by.x = c("DBN"),
                  by.y = c("DBN"),
                  all.y=TRUE)

demosnapshot_SAT <- demosnapshot_SAT[complete.cases(demosnapshot_SAT),]
```


```{r}
snapshot_5n_SAT <- merge(demosnapshot, demosnapshot_SAT, 
                  by.x = c("DBN"),
                  by.y = c("DBN"),
                  all.y=TRUE)

# Need to discard extra columns
snapshot_5n_SAT <- snapshot_5n_SAT[,1:14]
```

```{r}
snapshot_5n_SAT %>% group_by("Year.x" = Year.x) %>% summarize(count=n())
```

```{r}
# Need to create procedure for new predictions employing Difference Equations
demosnapshot_MODEL <- data.frame( "DBN"= 0, "School.Name" = "Need", "Year" = 0, "Total.Enrollment" = 0,
                                  "X..Female.1" = 0, "X..Male.1" = 0, "X..Asian.1" = 0, "X..Black.1" = 0,
                                  "X..Hispanic.1" = 0, "X..Other.1" = 0, "X..White.1" = 0, 
                                  "X..Students.with.Disabilities.1" = 0, "X..English.Language.Learners.1" = 0,  "X..Poverty.1" = 0  )

# Need Schoo Name as character
demosnapshot_MODEL$School.Name <- as.character(demosnapshot_MODEL$School.Name)
# Need to define n
n <- c(0,1,2,3,4)
# Initializing b and i
b <- 0
i <- 1

```



```{r}

for (DBN in unique(snapshot_5n_SAT$DBN)){

  # Need to run columns to the right
  for (c in 4:14){

      # Extracting pn from each column
      pn <- snapshot_5n_SAT[snapshot_5n_SAT$DBN == DBN, c]
      df <- data.frame(n,pn)
      
      # Need to calculate Delta_n
      df <- data.table(df)
      df[ , Delta := shift(pn, type ="lead") - pn]
      
      # Calculating slope k employing linear model in r with intersection
      lm_model <- lm(Delta ~ n , data = df)
      
      # Extracting slope coeficient from linear model
      k <- lm_model$coefficients[[2]]
      b <- lm_model$coefficients[[1]]

      # Need to sort and take top 2 highest values
      #top2 <- sort(pn)
      #M <- (top2[4] + top2[5])/2
        
      # Need to find new k
      #df$Delta[5] <- 0
      #k1 <- df$Delta / ((M - df$pn) * df$pn)
      #k1 <- sort(k[1:4])
      #k1 <- k1[4]      
      
      # Need to run each row in the column
      df <- data.frame(n = df$n, Delta = df$Delta, predict = df$pn[1], pn = pn)
      pred <- round(pn + k * n + b, 1)
      # Need to calculate new modeled predictions
      for (r in 2:6){
        df[r,1] <- r - 1
        pnn <- df[r-1,3]
        
        # Linear model
        
        df[r,3] <- round(pnn + k * r + b, 1)

        # Refined by finding max approximation
        #df[r,3] <- round(pnn + k * (M - pnn) * pnn, 1)

      }
      
      # Need to create new data frame with new predicted value
      demosnapshot_MODEL[i,c] <- df[6,3]
      
  }
  demosnapshot_MODEL[i,1] <- DBN
  demosnapshot_MODEL[i,2] <- snapshot_5n_SAT[snapshot_5n_SAT$DBN == DBN, 2][1]
  demosnapshot_MODEL[i,3] <-"2016-17"
  
  i <-  i + 1
}

```

```{r}
# Need to take care of constrains; that is...
# a) No negative numbers, the minimum is 0 on all variables
# b) The minimum percentage is 0 and maximum percentage is 100

demosnapshot_MODEL$Total.Enrollment <- round(demosnapshot_MODEL$Total.Enrollment,0)
demosnapshot_MODEL$Total.Enrollment[demosnapshot_MODEL$Total.Enrollment < 0] <- 0


for (c in 5:14){
     demosnapshot_MODEL[,c][demosnapshot_MODEL[,c] < 0] <- 0 
     demosnapshot_MODEL[,c][demosnapshot_MODEL[,c] > 100] <- 100 
}
      
```


# Prediction process for SAT 2016-17

```{r}
# Running linear models on predicted data
predict_R_SAT <- predict(Rglm.SAT_STEP, newdata=demosnapshot_MODEL, type="response")
predict_W_SAT <- predict(Wglm.SAT_STEP, newdata=demosnapshot_MODEL, type="response")
predict_M_SAT <- predict(Mglm.SAT_STEP, newdata=demosnapshot_MODEL, type="response")

```

```{r}
par(mfrow=c(3,2))
hist(predict_R_SAT)
hist(SAT$SAT.Critical.Reading.Avg..Score)
hist(predict_W_SAT)
hist(SAT$SAT.Writing.Avg..Score)
hist(predict_M_SAT)
hist(SAT$SAT.Math.Avg..Score)

```


\newpage
# CONCLUSIONS

In this section we need to add conclusions...




\newpage
# REFERENCES

Need to add references...

